---
title: "Complete Individual Maps"
format: html
editor: visual
execute: 
  echo: false
---

# Results

```{r}
#| warning: false
#| include: false
library(tidycensus)
library(tidyverse)
library(ggplot2)
library(gganimate)
library(gifski)
library(sf)
```

```{r}
#| warning: false
#| include: false
prefixAg <- read_csv("prefixAg.csv", col_types = cols(ddodyear = col_character(),PUMA = col_character()))
mutateAg <- read_csv("mutateAg.csv", col_types = cols(ddodyear = col_character(), PUMA = col_character(), newPUMA = col_character()))
```

```{r}
#| warning: false
#| include: false
apiKey="d892b83d5d69d3c30133cbe104ba890201cecb27"
census_api_key(apiKey, install = TRUE, overwrite=TRUE)
```

```{r}
#| warning: false
#| include: false
popData2013 <- get_acs(
    geography = "public use microdata area",
    variables = c(total_population = "B01003_001"),
    state = "OR",
    year = 2013,
    survey = "acs5",
    geometry = TRUE, output = "wide") 

popData2014 <- get_acs(
    geography = "public use microdata area",
    variables = c(total_population = "B01003_001"),
    state = "OR",
    year = 2014,
    survey = "acs5",
    geometry = TRUE, output = "wide") 

popData2015 <- get_acs(
    geography = "public use microdata area",
    variables = c(total_population = "B01003_001"),
    state = "OR",
    year = 2015,
    survey = "acs5",
    geometry = TRUE, output = "wide") 

popData2016 <- get_acs(
    geography = "public use microdata area",
    variables = c(total_population = "B01003_001"),
    state = "OR",
    year = 2016,
    survey = "acs5",
    geometry = TRUE, output = "wide") 

popData2017 <- get_acs(
    geography = "public use microdata area",
    variables = c(total_population = "B01003_001"),
    state = "OR",
    year = 2017,
    survey = "acs5",
    geometry = TRUE, output = "wide") 

popData2018 <- get_acs(
    geography = "public use microdata area",
    variables = c(total_population = "B01003_001"),
    state = "OR",
    year = 2018,
    survey = "acs5",
    geometry = TRUE, output = "wide") 

popData2019 <- get_acs(
    geography = "public use microdata area",
    variables = c(total_population = "B01003_001"),
    state = "OR",
    year = 2019,
    survey = "acs5",
    geometry = TRUE, output = "wide") 

popData2020 <- get_acs(
    geography = "public use microdata area",
    variables = c(total_population = "B01003_001"),
    state = "OR",
    year = 2020,
    survey = "acs5",
    geometry = TRUE, output = "wide") 

popData2021 <- get_acs(
    geography = "public use microdata area",
    variables = c(total_population = "B01003_001"),
    state = "OR",
    year = 2021,
    survey = "acs5",
    geometry = TRUE, output = "wide") 

popData2022 <- get_acs(
    geography = "public use microdata area",
    variables = c(total_population = "B01003_001"),
    state = "OR",
    year = 2022,
    survey = "acs5",
    geometry = TRUE, output = "wide") 
```

```{r}
#| warning: false
#| include: false
Ag13 <- mutateAg|>
  filter(ddodyear == 2013)

CancerRates2013 <- popData2013 |>
  right_join(Ag13, by = c("GEOID" = "newPUMA")) |>
  mutate(rate_per_100k = (totals / total_populationE) * 100000)


Ag14 <- mutateAg|>
  filter(ddodyear == 2014)

CancerRates2014 <- popData2014 |>
  right_join(Ag14, by = c("GEOID" = "newPUMA")) |>
  mutate(rate_per_100k = (totals / total_populationE) * 100000)


Ag15 <- mutateAg|>
  filter(ddodyear == 2015)

CancerRates2015 <- popData2015 |>
  right_join(Ag15, by = c("GEOID" = "newPUMA")) |>
  mutate(rate_per_100k = (totals / total_populationE) * 100000)

Ag16 <- mutateAg|>
  filter(ddodyear == 2016)

CancerRates2016 <- popData2016 |>
  right_join(Ag16, by = c("GEOID" = "newPUMA")) |>
  mutate(rate_per_100k = (totals / total_populationE) * 100000)

Ag17 <- mutateAg|>
  filter(ddodyear == 2017)

CancerRates2017 <- popData2017 |>
  right_join(Ag17, by = c("GEOID" = "newPUMA")) |>
  mutate(rate_per_100k = (totals / total_populationE) * 100000)

Ag18 <- mutateAg|>
  filter(ddodyear == 2018)

CancerRates2018 <- popData2018 |>
  right_join(Ag18, by = c("GEOID" = "newPUMA")) |>
  mutate(rate_per_100k = (totals / total_populationE) * 100000)

Ag18 <- mutateAg|>
  filter(ddodyear == 2018)

CancerRates2018 <- popData2018 |>
  right_join(Ag18, by = c("GEOID" = "newPUMA")) |>
  mutate(rate_per_100k = (totals / total_populationE) * 100000)

Ag19 <- mutateAg|>
  filter(ddodyear == 2019)

CancerRates2019 <- popData2019 |>
  right_join(Ag19, by = c("GEOID" = "newPUMA")) |>
  mutate(rate_per_100k = (totals / total_populationE) * 100000)

Ag20 <- mutateAg|>
  filter(ddodyear == 2020)

CancerRates2020 <- popData2020 |>
  right_join(Ag20, by = c("GEOID" = "newPUMA")) |>
  mutate(rate_per_100k = (totals / total_populationE) * 100000)

Ag21 <- mutateAg|>
  filter(ddodyear == 2021)

CancerRates2021 <- popData2021 |>
  right_join(Ag21, by = c("GEOID" = "newPUMA")) |>
  mutate(rate_per_100k = (totals / total_populationE) * 100000)

Ag22 <- prefixAg|>
  filter(ddodyear == 2022)

CancerRates2022 <- popData2022 |>
  right_join(Ag22, by = c("GEOID" = "PUMA")) |>
  mutate(rate_per_100k = (totals / total_populationE) * 100000)

```

```{r}
CancerRates <- bind_rows(CancerRates2013, CancerRates2014, CancerRates2015, CancerRates2016, CancerRates2017, CancerRates2018, CancerRates2019, CancerRates2020, CancerRates2021, CancerRates2022)

CancerRatesLung <- CancerRates %>% filter(diseaseSite == "Lung")

CancerRatesPancreas <- CancerRates %>% filter(diseaseSite == "Pancreas")

CancerRatesBreast <- CancerRates %>% filter(diseaseSite == "Breast")

#| echo: false
#| include: false
```

## Maps

```{r}
ggplot(CancerRatesLung) +
    geom_sf(mapping=aes(fill = rate_per_100k)) +
    scale_fill_viridis_c(
      name = "Lung Cancer Deaths/100k residents",
      limits = c(0, 50)  # Adjust these limits based on your data

    ) +
    theme_void() + #removes axes 
    labs(title= "Lung Cancer Deaths/100k residents, 2013 - 2022: {current_frame}") + transition_manual(ddodyear) 

```

```{r}
ggplot(CancerRatesPancreas) +
    geom_sf(mapping=aes(fill = rate_per_100k)) +
    scale_fill_viridis_c(
      name = "Pancreas Cancer Deaths/100k residents",
      limits = c(0, 20)  # Adjust these limits based on your data

    ) +
    theme_void() + #removes axes 
    labs(title= "Estimated pancreas cancer deaths per 100k residents, 2013 - 2022: {current_frame}") + transition_manual(ddodyear) 
```

```{r}
ggplot(CancerRatesBreast) +
    geom_sf(mapping=aes(fill = rate_per_100k)) +
    scale_fill_viridis_c(
      name = "Breast Cancer Deaths/100k residents",
      limits = c(0, 25)  # Adjust these limits based on your data

    ) +
    theme_void() + #removes axes 
    labs(title= "Estimated breast cancer deaths per 100k residents, 2013 - 2022: {current_frame}") + transition_manual(ddodyear) 
```

### Results of Statistical Tests

One way ANOVA tests revealed no significant differences in the average mortality rates of breast cancer or pancreatic cancer based on geography (p \> 0.05).

Geographic area did have an impact on lung cancer mortality (p = 0.000004). Tukey's HSD pairwise comparisons revealed that East Portland had significantly higher lung cancer mortality rates compared to the North and Northeast (p = 0.005), Central East (p = 0.0045), and Northwest & Southwest (p = 0.0000012) areas. Additionally, Northwest & Southwest Portland had significantly lower rates than the areas of East Portland (p = 0.0000012), Southeast Portland (p = 0.0015), and Gresham & Troutdale cities (p = 0.0047).
